Draft Article

### 2019-08-29 article.tex
started literature review

I posit that the universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives. On balance, states derive a benefit from preserving a system in which the consequences for espionage are minimized, so they refrain from imposing particularly harsh punishments for its practice; even when an individual incident is especially damaging to a state's security, if it falls within the traditionally accepted boundaries of an espionage operation, then it will likely go unpunished. In the context of the OPM hack, that means that US policymakers decided that they could not impose harsh penalties on the Chinese because it was more valuable to them that the US retain its ability to practice cyber-espionage with relative impunity in the future. This basic calculation has remained consistent since the advent of the modern peacetime intelligence service, perhaps even the nation-state itself.

This article seeks to establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it. Through this framework, I explain what constitutes traditional espionage and why rational states might calculate that it is within their best interests to continue tolerating it. I then examine two cases in which new technical means of gathering intelligence were introduced during the Cold War: aerial photography (spy planes) and satellite photography (spy satellites). In both cases, the norms of traditional espionage ultimately prevailed, minimizing the diplomatic consequences for their use as long as it was in service of traditional intelligence goals. Through process tracing, I demonstrate how the justifications that policymakers employed for their use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for why an accord tacitly permitting the use of espionage could be mutually strategic.

\subsection{Literature base}
This article makes three main contributions to the academic literature on espionage. The first is that it advances a theory for why espionage continues to retain its liminal status in the international legal system, where it is disavowed but allowed. Public discourse on this subject often uses terms like \enquote{rules of the road} or \enquote{gentleman's agreement} to describe the understanding between governments that espionage should fall within certain parameters. By employing both realist and constructivist schools of IR, this article develops rigorous explanations for why rational states, whether they be security-seeking and greedy, might cooperate to preserve this dynamic.

Previous academic literature on espionage and diplomacy has treated espionage as \emph{sui generis}, a legal outlier whose norms are entirely endogenous to the insular intelligence community.

Rather than explain the constituent practices of espionage as historically contingent,

Typically those consequences are minimal, rarely amounting to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it. Most importantly, those consequences never rise to the level where they impose additional costs on the state that attempted the espionage in the first place; in the rare cases when costs \emph{are} imposed for a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again. Espionage retains a special status within the international community: the activity itself is expressly forbidden, but every state tacitly acknowledges that allies and adversaries alike are guaranteed to engage in it.

\section{Existing research on espionage}
Previous scholarly work examines the curious liminal space in which espionage exists, but not the rationale that states have for perpetuating it.

\section{Cyber-espionage as espionage}
At first glance this observation might seem almost tautological. Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage. That formulation begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage, what purpose that distinction servers, or whether it is prudent to continue making it.

>>>>

I posit that the universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives. On balance, states derive a benefit from preserving a system in which the consequences for espionage are minimized, so they refrain from imposing particularly harsh punishments for its practice; even when an individual incident is especially damaging to a state's security, if it falls within the traditionally accepted boundaries of an espionage operation, then it will likely go unpunished. In the case of the OPM hack, US policymakers decided that they could not impose harsh penalties on the Chinese because it was more valuable to them that the US retain its ability to practice cyber-espionage with relative impunity in the future. This basic calculation has remained consistent since the advent of the modern peacetime intelligence service, perhaps even the nation-state itself.

This article seeks to establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it. Through this framework, I explain what constitutes traditional espionage and why rational states might calculate that it is within their best interests to continue tolerating it. I then examine two cases in which new technical means of gathering intelligence were introduced during the Cold War: aerial photography (spy planes) and satellite photography (spy satellites). In both cases, the norms of traditional espionage ultimately prevailed, minimizing the diplomatic consequences for their use as long as it was in service of traditional intelligence goals. Through process tracing, I demonstrate how the justifications that policymakers employed for their use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for why a tacit accord permitting the use of espionage could be mutually strategic.

\subsection{Existing literature}
Before continuing, it is necessary to properly define what I mean by espionage norms. While there are a number of practices that make espionage unique in IR, the norms examined in this article are the ones that govern the diplomatic protocols for intelligence. What consequences does a state impose when it uncovers an espionage operation run by another state? Espionage norms dictate that the consequences remain minimal. Rarely does the response amount to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it. Most importantly, those consequences never rise to the level where they impose additional costs on the state that attempted the espionage in the first place; in the rare cases when costs \emph{are} imposed for a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.

This article makes three main contributions to the academic literature on espionage. The first is that it advances a novel theory for why espionage retains its often contradictory status in the legal system. In the absence of any significant treaties regarding espionage, scholars must develop their own theories about its place in customary international law.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage. In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments. See \cite{wright_essays_1962}} Espionage is illegal everywhere, so how can every state acknowledge that allies and adversaries alike are guaranteed to engage in it?\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is. If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero. This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} There is no legally consistent way to resolve that, so scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact. Jared Beim's Comment determines that there while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so the prohibition on espionage merits further enforcement.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe. For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that his conclusion will have little impact on the practice of espionage, where, at best, \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[p.~672]{beim_enforcing_2018} A. John Radsan takes a more philosophical approach. Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{Beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage all come to the same conclusion: it would be almost impossible to argue that espionage is legal, but it will likely continue to be downplayed because the way it's been for decades. This literature cannot explain why espionage norms require this toleration.

% espionage is an inherently unique category of IR, whose norms are endogenous to the insular intelligence community and therefore bear little relation to other legal or strategic considerations.

Another branch of

I break from this literature by arguing that far from being an irreplicable, historically contingent set of practices, espionage norms are actually preserved through an ongoing set of strategic decisions. Each time a new espionage technology is made available, policymakers are given a fresh chance to decide whether or not the norms of espionage out to extend to its use. Critically, the research will show that they were not obligated to do so by any pre-existing norms. When the US first introduced photographic satellites, for instance, it was not clear whether the Soviet Union would treat them as spies or as weapons---there were indications in both directions. Ultimately, the infrastructure of what we now call \enquote{spy satellites} was developed, and the norms of espionage prevailed.

While not necessarily indicative of leading scholarship, it is revealing that popular discourse about espionage often uses terms like \enquote{rules of the game} or \enquote{gentleman's agreement} to describe the understanding between governments that espionage should fall within certain parameters. That these informal rules exist in some capacity is beyond a doubt.

% espionage as deescalataory

the second contribution is the role of espionage in deescalating conflict

Because I argue that espionage norms influence cyber policy, it is worth mentioning scholarship on the role that espionage, outside of cyberspace, plays in the international system today. The intersection of espionage and international relations has been studied before---beginning, in the modern context, as far back as 1962---but usually in a historical context, not a political science one. The role of espionage in deescalating conflict is typically relegated to accounts of moments where Great People properly interpreted---or fatally misunderstood---the information that their intelligence operations provided them. Chapters 3 and 4 will primarily focus on decisions made by Eisenhower and Khrushchev, so to some extent I am guilty of that as well. Nonetheless, I will try to highlight the way in which both of those leaders set precedents designed to outlast them, across a variety of intelligence methods. I hope that this thesis starts the conversation about how espionage norms operate to deescalate conflict independent of the leaders that promote them.

By employing both realist and constructivist schools of IR, this article develops rigorous explanations for why rational states, whether they be security-seeking and greedy, might cooperate to preserve this dynamic.

% incorporating cyber espionage into the intelligence literature, rather than the cyebr literature
% espionage first, cyber second
At first glance this observation might seem almost tautological. Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage. That formulation begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage, what purpose that distinction servers, or whether it is prudent to continue making it.
